Learning \strips\ action models with classical planning

### 2017-11-01 doc/icaps-18/strips-learning.tex
updated page 5 and merged with last results by Diego

Figure~\ref{fig:plan} shows a classical plan for solving a learning task of the $\Lambda$ kind. This classical plan programs and validates the operator schema {\tt\small stack} from the blocksworld (previously specified operator schemes for $pickup$, $putdown$ and $unstack$ are given), using the plan $\pi_1$ and label $\sigma_1$ shown in Figure~\ref{fig:lexample}. The plan steps $[0,8]$ are the actions for programming the preconditions of the {\tt\small stack} operator, steps $[9,13]$ are the actions for programming the operator effects and steps $[14,22]$ are actions for validating the programmed operators. The details of the compilation are explained in the remaining of the section.

@@

\section{Constraining the learning hypothesis space with additional input knowledge}
Here we show that further input knowledge can be used to constrain the space of the possible action models and make the learning of \strips action models more practicable.

\subsection{Example plans}
Here  we extend our compilation to address the learning scenario where, a set of plans is available. Given a learning task $\Lambda=\tup{\Psi,\Pi,\Sigma}$, the compilation outputs a classical planning task $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I_{\Lambda},G_{\Lambda}}$ that extends $P_{\Lambda}'$ as follows:
\begin{itemize}
\item $F_{\Lambda}$ extends $F_{\Lambda'}$ with $F_{\Pi}=\{plan(name(\xi),\Omega^{ar(\xi)},j)\}$, the fluents to code the $j$ steps of plans in $\Pi$, where $F_{\pi_t}\subseteq F_{\Pi}$ encodes $\pi_t\in \Pi$. Fluents $at_j$ and $next_{j,j_2}$, {\small $1\leq j<j2\leq n$}, are also added to represent the current and next plan steps.
\item $I_{\Lambda}$ is extended with the fluents from $F_{\Pi}$ that encode the plan $\pi_1\in \Pi$, the fluents $at_1$ and $\{next_{j,j_2}\}$, {\small $1\leq j<j2\leq n$}, for indicating where to start validating the programmed action model. Goals are the same as in the previous compilation $G_{\Lambda}=G_{\Lambda}'=\bigcup_{1\leq t\leq \tau}\{test_t\}$.
\item With respect to $A_{\Lambda}$.
\begin{enumerate}
\item The actions for programming the preconditions/effects of a given operator schema $\xi\in\Xi$ are the same.
\item The actions for applying an already programmed operator have an extra precondition $f\in F_{\Pi}$, that encodes the current plan step, and extra conditional effects $\{at_{j}\}\rhd\{\neg at_{j},at_{j+1}\}_{\forall j\in [1,n]}$ for advancing to the next plan step. This mechanism forces that these actions are only applied as in the example plans.
\item The actions for updating the active test have an extra precondition, $at_{|\pi_t|}$, to indicate that the current plan $\pi_t$ was fully executed and extra conditional effects to load the next plan $\pi_{t+1}$:
\begin{small}
\begin{align*}
&\{f\}\rhd\{\neg f\}_{f\in F_{\pi_t}}, \{\emptyset\}\rhd\{f\}_{f\in F_{\pi_t+1}},\{\emptyset\}\rhd\{\neg at_{|\pi_t|},at_1\}.
\end{align*}
\end{small}
\end{enumerate}
\end{itemize}

\subsection{Previously specified action models}
Our compilation does not require to start learning \strips action models from scratch and accepts partially specified operator schema where some preconditions and effects are a priori known.

The known preconditions and effects are encoded as fluents $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$ that are part of the initial state $I_{\Lambda'}$. The corresponding programming actions ($\mathsf{programPre_{f,\xi}}$ and $\mathsf{programEff_{f,\xi}}$) become unnecessary and are removed from $A_{\Lambda'}$ making the classical planning task $P_{\Lambda'}$ easier to be solved. To illustrate this, the classical plan of Figure~\ref{fig:plan} is a solution to a learning task for getting the blocksworld action model where the schemes for {\tt\small pickup}, {\tt\small putdown} and {\tt\small unstack} are known in advance.

In the extreme, when a fully specified \strips action model is given, the compilation validates whether an observed plan follows the given model. In this case, if a solution plan is found to $P_{\Lambda'}$, it means that the given \strips action model is {\em valid} for the given set of examples. If $P_{\Lambda'}$ is unsolvable it means that the given \strips action model is invalid since it is not compliant with all the given examples. Tools for plan validation like VAL~\cite{howey2004val} could also be used at this point.

\subsection{Static predicates}
A {\em static predicate} $p \in \Psi$ is a predicate that does not appear in the effects of any action schema~\cite{fox:TIM:JAIR1998}. Therefore, one can get rid of the mechanism for programming static predicates as action effects keeping the compilation complete. Formally, given a static predicate $p$:
\begin{itemize}
\item Fluents $del_f(\xi)$ and $add_f(\xi)$, such that $f\in F_v$ is an instantiation of the static predicate $p$ in the set of {\em variable} objects $\Omega_v$, can be discarded for every $\xi\in\Xi$.
\item Actions $\mathsf{programEff_{f,\xi}}$ (s.t. $f\in F_v$ is an instantiation of $p$ in $\Omega_v$) can also be discarded for every $\xi\in\Xi$.
\end{itemize}

Static predicates can also be used to constrain the space of possible preconditions by looking at the given labels. If a precondition $f\in F_v$ (s.t. $f\in F_v$ is an instantiation of a static predicate in $\Omega_v$) is not possible according to $\Sigma$, then the fluents $pre_f(\xi)$ and actions $\mathsf{programPre_{f,\xi}}$ can be discarded for every $\xi\in\Xi$. For instance in the {\em zenotravel} domain fluents $pre\_next\_board\_v1\_v1$, $pre\_next\_debark\_v1\_v1$, $pre\_next\_fly\_v1\_v1$, $pre\_next\_zoom\_v1\_v1$, $pre\_next\_refuel\_v1\_v1$ can be discarded (as well as their corresponding programming actions) because a precondition {\tt\small(next ?v1 ?v1 - flevel)} cannot be compliant with any state in $\Sigma$.

Likewise, static predicates can constrain the space of possible preconditions looking at the given example plans. In this case, fluents $pre_f(\xi)$ and actions $\mathsf{programPre_{f,\xi}}$ are discarded for every $\xi\in\Xi$ if a precondition $f\in F_v$ (s.t. $f\in F_v$ is an instantiation of a static predicate in $\Omega_v$) is not possible according to $\Pi$. Back to the {\em zenotravel} domain, if a example plan $\pi_t\in \Pi$ contains the action {\tt\small (fly plane1 city2 city0 fl3 fl2)} and the corresponding label $\sigma_t\in\Sigma$ contains the static fluent {\tt\small (next fl2 fl3)} but it does not include {\tt\small (next fl2 fl2)}, {\tt\small (next fl3 fl3)} or {\tt\small (next fl3 fl2)} it means that the only possible precondition including the static predicate is $pre\_next\_fly\_v1\_v2$.

In the evaluation of our approach we do not assume that the set of static predicates is given but compute a set of candidates of static predicates from the input examples to the learning task.

@@

>>>>

\section{Constraining the learning hypothesis space with additional input knowledge}
Here we show that further input knowledge can be used to constrain the space of the possible action models and make the learning of \strips\ action models more practicable.

\subsection{Example plans}
Here  we extend our compilation to address the learning scenario where, a set of plans is available. Given a learning task $\Lambda'=\tup{\Psi,\Pi,\Sigma}$, the compilation outputs a classical planning task $P_{\Lambda'}=\tup{F_{\Lambda'},A_{\Lambda'},I_{\Lambda'},G_{\Lambda'}}$ that extends $P_{\Lambda}$ as follows:
\begin{itemize}
\item $F_{\Lambda'}$ extends $F_{\Lambda}$ with $F_{\Pi}=\{plan(name(\xi),\Omega^{ar(\xi)},j)\}$, the fluents to code the $j$ steps of plans in $\Pi$, where $F_{\pi_t}\subseteq F_{\Pi}$ encodes $\pi_t\in \Pi$. Fluents $at_j$ and $next_{j,j_2}$, {\small $1\leq j<j2\leq n$}, are also added to represent the current and next plan steps.
\item $I_{\Lambda'}$ is extended with the fluents from $F_{\Pi}$ that encode the plan $\pi_1\in \Pi$, the fluents $at_1$ and $\{next_{j,j_2}\}$, {\small $1\leq j<j2\leq n$}, for indicating where to start validating the programmed action model. Goals are the same as in the previous compilation $G_{\Lambda'}=G_{\Lambda}=\bigcup_{1\leq t\leq \tau}\{test_t\}$.
\item With respect to $A_{\Lambda'}$.
\begin{enumerate}
\item The actions for programming the preconditions/effects of a given operator schema $\xi\in\Xi$ are the same.
\item The actions for applying an already programmed operator have an extra precondition $f\in F_{\Pi}$, that encodes the current plan step, and extra conditional effects $\{at_{j}\}\rhd\{\neg at_{j},at_{j+1}\}_{\forall j\in [1,n]}$ for advancing to the next plan step. This mechanism forces that these actions are only applied as in the example plans.
\item The actions for updating the active test have an extra precondition, $at_{|\pi_t|}$, to indicate that the current plan $\pi_t$ was fully executed and extra conditional effects to unload plan $\pi_{t}$ and load the next plan $\pi_{t+1}$:
\begin{small}
\begin{align*}
&\{f\}\rhd\{\neg f\}_{f\in F_{\pi_t}}, \{\emptyset\}\rhd\{f\}_{f\in F_{\pi_t+1}},\{\emptyset\}\rhd\{\neg at_{|\pi_t|},at_1\}.
\end{align*}
\end{small}
\end{enumerate}
\end{itemize}

\subsection{Previously specified action models}
Our compilation does not require to start learning \strips\ action models from scratch and accepts partially specified operator schema where some preconditions and effects are a priori known.

The known preconditions and effects are encoded as fluents $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$ set to true at the initial state $I_{\Lambda'}$. The corresponding programming actions ($\mathsf{programPre_{f,\xi}}$ and $\mathsf{programEff_{f,\xi}}$) become unnecessary and are removed from $A_{\Lambda'}$ making the classical planning task $P_{\Lambda'}$ easier to be solved.

To illustrate this, the classical plan of Figure~\ref{fig:plan} is a solution to a learning task for getting the blocksworld action model where the schemes for {\tt\small pickup}, {\tt\small putdown} and {\tt\small unstack} are known in advance. This classical plan programs and validates the operator schema {\tt\small stack} from the blocksworld (previously specified operator schemes for $pickup$, $putdown$ and $unstack$ are given), using the plan $\pi_1$ and label $\sigma_1$ shown in Figure~\ref{fig:lexample}. The plan steps $[0,8]$ are the actions for programming the preconditions of the {\tt\small stack} operator, steps $[9,13]$ are the actions for programming the operator effects and steps $[14,22]$ are actions for validating the programmed operators.

In the extreme, when a fully specified \strips\ action model is given, the compilation validates whether an observed plan follows the given model. In this case, if a solution plan is found to $P_{\Lambda'}$, it means that the given \strips\ action model is {\em valid} for the given set of examples. If $P_{\Lambda'}$ is unsolvable it means that the given \strips\ action model is invalid since it is not compliant with all the given examples. Tools for plan validation like VAL~\cite{howey2004val} could also be used at this point.

@@

\subsection{Static predicates}
A {\em static predicate} $p \in \Psi$ is a predicate that does not appear in the effects of any action schema~\cite{fox:TIM:JAIR1998}. Therefore, one can get rid of the mechanism for programming these predicates as the effect of any action schema while keeping the compilation complete. Formally, given a static predicate $p$:
\begin{itemize}
\item Fluents $del_f(\xi)$ and $add_f(\xi)$, such that $f\in F_v$ is an instantiation of the static predicate $p$ in the set of {\em variable} objects $\Omega_v$, can be discarded for every $\xi\in\Xi$.
\item Actions $\mathsf{programEff_{f,\xi}}$ (s.t. $f\in F_v$ is an instantiation of $p$ in $\Omega_v$) can also be discarded for every $\xi\in\Xi$.
\end{itemize}

Static predicates can be used to constrain the space of possible preconditions by looking at the given set of labels $\Sigma$. In particular one can assume that if a precondition $f\in F_v$ (s.t. $f\in F_v$ is an instantiation of a static predicate in $\Omega_v$) is not compliant with the labels in $\Sigma$ it means that is not possible and then, fluents $pre_f(\xi)$ and actions $\mathsf{programPre_{f,\xi}}$ can be discarded for every $\xi\in\Xi$. For instance in the {\em zenotravel} domain fluents $pre\_next\_board\_v1\_v1$, $pre\_next\_debark\_v1\_v1$, $pre\_next\_fly\_v1\_v1$, $pre\_next\_zoom\_v1\_v1$, $pre\_next\_refuel\_v1\_v1$ can be discarded (as well as their corresponding programming actions) because a precondition {\tt\small(next ?v1 ?v1 - flevel)} cannot be compliant with any state in $\Sigma$.

Likewise, static predicates can constrain the space of possible preconditions looking at the given set of example plans. In this case, fluents $pre_f(\xi)$ and actions $\mathsf{programPre_{f,\xi}}$ are discarded for every $\xi\in\Xi$ if a precondition $f\in F_v$ (s.t. $f\in F_v$ is an instantiation of a static predicate in $\Omega_v$) is not possible according to $\Pi$. Back to the {\em zenotravel} domain, if a example plan $\pi_t\in \Pi$ contains the action {\tt\small (fly plane1 city2 city0 fl3 fl2)} and the corresponding label $\sigma_t\in\Sigma$ contains the static literal {\tt\small (next fl2 fl3)} but it does not contain {\tt\small (next fl2 fl2)}, {\tt\small (next fl3 fl3)} or {\tt\small (next fl3 fl2)} it means that the only possible precondition including the static predicate is $pre\_next\_fly\_v1\_v2$.

In the evaluation of our approach we do not assume that the set of static predicates is given but compute a set of {\em potential} static predicates from the input examples to the learning task.

@@

\newpage
